One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2015-01-01 paper/model.tex
major restructuring and rewrite in response to report

\section{Governing equations and numerical models}\label{model} 
We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
orbiting a central star of mass $M_*$. We will mainly examine 2D
(or razor-thin) discs in favour of numerical 
resolution, but have also carried out some 3D disc
simulations. Hence, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the 
fluid velocity (the angular velocity being $\Omega \equiv v_\phi/R$), 
$p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed $c_s$ is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient (since, for an ideal gas the
temperature is proportional to $c_s^2$). The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
 disc with constant $H/R$.

@@

@@

\section{Linear density waves}\label{wkb}
In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
stability of disc equilibria described 
above. However, it is useful to  interpret simulation results
using linear density theory. We list here the main definitions and
results from 2D linear theory, taken from \cite{lin11b} and
\cite{shu91}.   In particular, we highlight a key feature of
locally isothermal discs that enables angular momentum exchange
between linear perturbations and the background disc.

In a linear analysis, one assumes a steady axisymmetric background state, such as
that defined in \S\ref{model}, which is then perturbed such that
\begin{align}  
  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
        m\phi\right)\right]}, 
\end{align}
and similarly for other variables, where $\sigma=\omega+\ii\gamma$ is
a complex frequency with $\omega$ being the real frequency, 
$\gamma$ the growth rate, and $m$ is an integer. Here, we take $m>0$ without
loss of generality.  
%Since the background is
%axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
 %   t\right)} = \Sigma_m$ as defined for diagnostics.

These linearised equations can be combined into a single
integro-differential equation eigenvalue problem. We defer a full
exploration of the linear problem to a future study. Here, we list some
general  properties of the linear perturbations that will be relevant
for describing simulation results.

\subsection{Global angular momentum conservation for linear
  perturbations} \label{global_cons}
It can be shown that linear perturbations with $\phi$-dependence in the form
$\exp{(\ii m\phi)}$ satisfies angular momentum conservation
in the form 
\begin{align}\label{angcons}
  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
\end{align}
\citep[e.g.][]{narayan87,ryu92,lin93b} where
\begin{align}\label{lin_ang_mom_cons}
  \jlin \equiv
  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
    m \Omega |\bm{\xi}|^2\right) 
\end{align}
is the angular momentum density of the linear disturbance (which may
be positive or negative), where $\bm{\xi}$ is the Lagrangian
displacement and $^*$ denotes complex conjugation, and $\bm{F}$ is the
vertically-integrated angular momentum flux consisting of a Reynolds
stress and a gravitational stress \citep{lin11b}. Explicit expressions
for $\bm{\xi}$ can be found in, e.g. \cite{papaloizou85}.

In Eq. \ref{angcons}, the background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the total angular momentum associated with
the perturbation is conserved, provided that there is no net angular
momentum flux. % and the perturbations are regular.

However, as noted in \cite{lin11b}, if there is an imposed
temperature gradient, as in the disc models we consider,
then $\tbg\neq 0$ in general, which corresponds to a local torque
exerted by the background disc on the perturbation. We outline the
derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.

The important consequence of the background torque is the possibility
of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
(negative) and the background torque $\tbg$ is positive (negative),
then the local angular momentum density of the linear disturbance
further increases (decreases). This implies the amplitude of the disturbance
may grow by exchanging angular momentum with the background
disc. In our disc models with $c_s^2\propto R^{-q}$, we expect the
growth rate of such an instability to scale with $q$.

Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
\omega/m$ of a neutral mode 
, Eq. \ref{dispersion} can be solved 
for $|k|$, 
\begin{align}\label{wavenumber}
  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
      Q^2(1-\nu^2)}\right], 
\end{align}
where 
% \begin{align}
%   k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
% \end{align}
% is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. In
Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
spirals.

\section{Numerical methods}\label{methods}
We adopt computational units such that
$G=M_*=R_0=1$. Time is measured  in the Keplerian orbital period at
the reference radius, $P_0\equiv  2\pi/\Omega_k(R_0)$.

@@

>>>>

\section{Governing equations}\label{model} 
We consider an inviscid fluid disc of mass $M_d$ 
orbiting a central star of mass $M_*$. We will mainly examine 2D
(or razor-thin) discs in favour of numerical 
resolution, but have also carried out some 3D disc
simulations. Hence, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the 
fluid velocity (the angular velocity being $\Omega \equiv v_\phi/R$), 
$p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed $c_s$ is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient (since, for an ideal gas the
temperature is proportional to $c_s^2$). The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
 disc with constant $H/R$.

@@

\section{Linear density waves}\label{wkb}
% In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
% stability of disc equilibria described 
% above. However, it is useful to  interpret simulation results
% using linear density theory. We list here the main definitions and
% results from 2D linear theory, taken from \cite{lin11b} and
% \cite{shu91}.  
{\bf We highlight a key feature of locally isothermal discs that
  enables angular momentum exchange between small disturbances and the
  background disc through an imposed radial temperature gradient. This
  conclusion results from the consideration of angular momentum
  conservation within the framework of linear perturbation theory. 
  For simplicity, we consider 2D discs, although generalisation to 3D is straight
  forward.  
}

In a linear analysis, one assumes a steady axisymmetric background state, % such as
% that defined in \S\ref{model},
which is then perturbed such that
\begin{align}  
  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
        m\phi\right)\right]}, 
\end{align}
and similarly for other variables, where $\sigma=\omega+\ii\gamma$ is
a complex frequency with $\omega$ being the real frequency, 
$\gamma$ the growth rate, and $m$ is an integer. Here, we take $m>0$ without
loss of generality.  
%Since the background is
%axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
 %   t\right)} = \Sigma_m$ as defined for diagnostics.

These linearised equations can be combined into a single
integro-differential equation eigenvalue problem. We defer a full
{\bf numerical} exploration of the linear problem to a future study. Here, we {\bf
  discuss} some
general properties of the linear perturbations. % that will be relevant
% for describing simulation results.

\subsection{Global angular momentum conservation for linear
  perturbations} \label{global_cons}
It can be shown that linear perturbations with $\phi$-dependence in the form
$\exp{(\ii m\phi)}$ satisfies angular momentum conservation
in the form 
\begin{align}\label{angcons}
  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
\end{align}
\citep[e.g.][]{narayan87,ryu92,lin93b} where
\begin{align}\label{lin_ang_mom_cons}
  \jlin \equiv
  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
    m \Omega |\bm{\xi}|^2\right) 
\end{align}
is the angular momentum density of the linear disturbance (which may
be positive or negative), $\bm{\xi}$ is the Lagrangian
displacement and $^*$ denotes complex conjugation, and $\bm{F}$ is the
vertically-integrated angular momentum flux consisting of a Reynolds
stress and a gravitational stress \citep{lin11b}. Explicit expressions
for $\bm{\xi}$ can be found in, e.g. \cite{papaloizou85}.

In Eq. \ref{angcons}, the background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. {\bf  We outline the
derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.}

In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the total angular momentum associated with
the perturbation is conserved, provided that there is no net angular
momentum flux. However, as noted in \cite{lin11b}, if there is an imposed
temperature gradient, as in the disc models we consider,
then $\tbg\neq 0$ in general, which corresponds to a local torque
exerted by the background disc on the perturbation.

The important consequence of the background torque is the possibility
of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
(negative) and the background torque $\tbg$ is {\bf also} positive (negative),
then the local angular momentum density of the linear disturbance
{\bf will} further increase (decrease) {\bf with time}. This implies the amplitude of the disturbance
may grow by exchanging angular momentum with the background
disc.%  In our disc models with $c_s^2\propto R^{-q}$, we expect the
% growth rate of such an instability to scale with $q$.

{\bf 
  It is simplest to demonstrate instability by determining  
  the sign of the angular momentum density 
  $\jlin$ and the background torque $T_\mathrm{BG}$ for perturbations
  in the local approximation, which we review next. 
  % depends on the
  % type of perturbation being considered. 
  We then show that low-frequency perturbations ($|\omega|\ll\Omega$)
  are unstable when there a negative radial 
  temperature gradient is imposed.  
}

\subsection{Angular momentum density of  non-axisymmetirc low-frequency modes}
{\bf We now specialise to low-frequency modes and first determine the sign of its
  angular momentum density.} From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
the angular momentum density associated with linear waves is
\begin{align}
  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right].  
\end{align}
For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
term $\omega|\bm{\xi}|^2$, {\bf we find}
\begin{align}
  \jlin &\simeq \frac{m\Sigma\Omega}{2}\left[-m|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right]\notag\\
  & = -\frac{m\Sigma\Omega}{2}\left[ (m-1)|\bm{\xi}|^2 +  |\xi_R + \ii\xi_\phi|^2\right].
%  & \leq 0.
\end{align}

Thus, low-frequency non-axisymmetric modes ($m\geq1$)

Thus, low-frequency $m=1$ modes generally have negative
angular momentum. These are the modes observed in our simulations. 
If the mode loses (positive) angular momentum 
to the background, then we can expect instability. We demonstrate
below how this is possible through a forced temperature
gradient.

%and setting $m=1$,
%\begin{align}
 % \jlin &\simeq \frac{\Sigma\Omega}{2}\left[-|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
  %    \xi_R^*\xi_\phi\right)\right]\notag\\
%  & = -\frac{\Sigma\Omega}{2}|\xi_R + \ii\xi_\phi|^2\notag\\
 % & \leq 0.
%\end{align}

Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
\omega/m$ of a neutral mode 
, Eq. \ref{dispersion} can be solved 
for $|k|$, 
\begin{align}\label{wavenumber}
%  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
 %     Q^2(1-\nu^2)}\right], 
|k| = k_c\left[1 \pm \sqrt{1 -
     Q^2(1-\nu^2)}\right], 
\end{align}
where 
 \begin{align}
   k_c \equiv \frac{\pi G \Sigma}{c_s^2}
 \end{align}
 is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. In
Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
spirals.

\subsection{Unstable interaction between low-frequency modes 
  and the background disc due to an imposed temperature gradient}
%The $m=1$ spiral that emerges in our simulations have low frequency
%($\Omega_p < \Omega$), grows slowly ($\gamma\ll\Omega_p$) and its
%spatial form is consistent local theory (\S\ref{fargo_m1}). 
%{\bf
%  We first regard a low-frequency, $m=1$ disturbance 
 % as a mode within the framework of local theory, satisfying the
 % % temperature gradient ignored in local theory
 % dispersion relation Eq. \ref{dispersion}. }
%To demonstrate possible
%instability, 
{\bf 
Here we show that the torque density acting on a low-frequency, $m=1$
local mode due to the background temperature gradient can be negative, which 
would enforce the mode, because it has negative angular momentum.  
}

% Note that
% we will not be assuming a self-gravitating disc until the end of this
% section. 
%We begin by recalling the expression for the torque density due to a
%locally isothermal equation of state,
%\begin{align*}
 % T_\mathrm{BG} = - \frac{m}{2}\frac{dc_s^2}{dR}\imag\left(\delta\Sigma_m\xi_R^*\right).   
%\end{align*}
The Eulerian surface density perturbation is given by
\begin{align}\label{den_pert}
  \delta\Sigma_m = -\nabla\cdot\left(\Sigma\bm{\xi}\right) 
  = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma \xi_R\right) - \frac{\ii m}{R}\Sigma\xi_\phi.  
\end{align}
We invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and assume $|kR|\gg m$ so that the second term on the right hand
side of Eq. \ref{den_pert} can be neglected. Then    
\begin{align}
  \delta\Sigma _m  \simeq -\ii k \Sigma \xi_R.
\end{align}
{\bf Inserting this into the expression for the background torque,
  Eq. \ref{baroclinic_torque}, we find}
\begin{align}
  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. 
\end{align}

This torque density is negative for trailing waves in discs with
a fixed temperature profile decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
respectively). Note that this conclusion does not rely on the
low-frequency approximation and holds for any $m$.

However, if the linear disturbance under consideration \emph{is} a
low-frequency $m=1$ mode, then it has negative angular
momentum. {\bf Then} if it is a trailing spiral and $dc_s^2/dR<0$, as is typical
in astrophysical discs, then $T_\mathrm{BG}<0$, and
the background disc applies a negative torque on the disturbance,
which further decreases its angular momentum. This suggests the {\bf
  mode} 
amplitude will grow.

Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadratic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
\begin{align}\label{theoretical_rate0}
  2\gamma \sim
  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
    \xi_\phi|^2}. 
%  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
\end{align}
In Appendix \ref{horizontal_displacements}, we argue $\xi_\phi \simeq
2\ii \xi_R$ for the modes of interest. Then 
%To further simplify this expression, we need to relate $\xi_R$ and
%$\xi_\phi$. 
% \begin{align}
%   2\gamma \sim
%   -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
%     + \ii\xi_\phi|^2}.
% \end{align}
% Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
% in the local approximation, we have
% \begin{align}
%   2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
%     \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
%   \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
% \end{align}
% where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
% inequality. 
%where we used $dc_s^2/dR <0$ and $k>0$ to obtain the inequality. 
for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{\Omega} \sim q h
  \left(kH\right)\Omega.  
\end{align}

{\bf It is important to note that this growth is
due to the temperature gradient and does not depend explicitly on
whether or not the disc is self-gravitating, i.e. destabilisation is
not due to self-gravity.}

For perturbations in a self-gravitating disc with characteristic
wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$, as seen in
our simulations, we have 
\begin{align}\label{theoretical_rate}
  \gamma \sim \frac{qh}{2Q}\Omega,
\end{align}
where we used $c_s/R\sim h\Omega$.

%For our fiducial disc models with
%$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega \sim O(10^{-2})$,
%consistent with simulation results. Specifically,
%Eq. \ref{theoretical_rate} indicates the growth rate should increase
%linearly with $q$ and $h$, as seen in the simulations
%(Fig. \ref{fargo_varq_growth}---\ref{fargo_varh_growth}).

\section{Numerical simulations}\label{methods}
%above discussion is generic for low-frequency non-axisymmetric modes,
%but illustrate with a concrete example
We adopt computational units such that
$G=M_*=R_0=1$. Time is measured  in the Keplerian orbital period at
the reference radius, $P_0\equiv  2\pi/\Omega_k(R_0)$.

@@

@@

% \begin{align}
%   T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
%         v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
%   & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
%         v_{Rm}}{\sbar}\right)^*\right]
% \end{align}
% where we replaced the radial Lagrangian displacement by the Eulerian
% radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
% The linearised momentum equations give
% \begin{align}
%   D\delta v_{Rm} =& 
%   \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
%     + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
%   &- \frac{2\ii
%     m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
%     \delta\Phi_m\right),
% \end{align}
% where $D\equiv \kappa^2 - \sbar^2$. 
% We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
% real, and using the local solution to the Poisson equation
% \begin{align}
%   \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
% \end{align}
% \citep{shu91}. Then the radial velocity perturbation becomes
% \begin{align}
%   \delta v_{Rm} =
%   \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
%       m\Omega}{R}\right),
% \end{align}
% so that 
% \begin{align}
%   \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
%   \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
%       m\Omega}{R\sbar^*}\right). 
% \end{align}
% For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
% taking the imaginary part of the above expression, the torque
% density is 
% \begin{align}\label{tbg_explicit}
%   T_\mathrm{BG} &=
%   \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\right].
% % &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
% %         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
% %       G \Sigma}{|k|} - c_s^2\right)\right]. 
% \end{align}
% We now use the local dispersion relation,
% Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
% k^2c_s^2$ in the above expression. This gives
% \begin{align}\label{tbg_simple}
%   T_\mathrm{BG} =
%   \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
% \end{align}

### 2015-01-01 paper/paper.tex
major restructuring and rewrite in response to report

\maketitle
\begin{abstract} 
  We present numerical simulations of astrophysical discs with an 
  imposed temperature profile and in which a sub-portion of it is
  self-gravitating. This might represent, for example, mass  
  built-up in a circumstellar or protoplanetary
  disc at the outer edge of a `dead zone', in which there
  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
  We conduct numerical gravito-hydrodynamic experiments in 2D using
  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
  discs we observe the development of a 
  spiral mode with azimuthal wavenumber $m=1$ and co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly stationary 
  one-armed spiral pattern in the self-gravitating region of the
  disc. The development of this one-armed 
  spiral does not require the movement of  
  the central star, as found in previous studies. Instead, we find it
  to be associated with the imposed 
  temperature gradient and the radial disc structure. 
  We apply results from linear theory and attribute the
  growth of this spiral mode to the non-conservation of angular momentum
  for non-axisymmetric linear perturbations in discs with a forced temperature
  gradient. This may lead to an unstable exchange of angular momentum
  between the spiral mode 
  and the background disc. However, because the interaction does
  not explicitly depend on self-gravity, we suggest it may also
  affect low-frequency one-armed oscillations in non-self-gravitating
  discs. 
\end{abstract}

>>>>

\maketitle
\begin{abstract} 
  {\bf We describe a new mechanism that leads to the destabilisation of
  one-armed spiral or eccentric modes in astrophysical discs with 
  an imposed radial temperature gradient.  We use linear density wave
  theory 
  to show that non-axisymmetric perturbations generally do not conserve
  their angular momentum in the presence of a forced temperature
  gradient. This implies an exchange of angular momentum
  between linear perturbations and the background disc. In particular, 
  when the disturbance is a low-frequency, one-armed
  trailing spiral and the disc temperature decreases outwards, this 
  interaction is unstable and leads to the growth of the spiral mode.  
  We demonstrate this phenomenon through numerical hydrodynamic
  simulations of locally isothermal discs in 2D using the FARGO
  code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
  radially structured discs with a self-gravitating sub-portion which remains
  stable in the absence of a temperature gradient. However,
  when a temperature gradient is imposed we observe exponential growth of an one-armed  
  spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a low-frequency  
  spiral pattern in the self-gravitating region of the disc.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by the imposed temperature gradient does 
 not explicitly depend on self-gravity, we suggest this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs. 
}
\end{abstract}

%This might represent, for example, mass  
 % built-up in a circumstellar or protoplanetary
 % disc at the outer edge of a `dead zone', in which there
  %is inefficient magneto-hydrodynamic turbulence to enable accretion.
